### 2024-11-22 Article/Conclusions.tex
minima

In questo caso, una introduzione meno brusca, ricapitolando un po' la struttura e soprattutto il fine della tesi

@@

The result we ended up with is then $\Reg(\EGG))$-adhesivity, where $\EGG$ is a category of functors which have $\Set$ as codomain category.
The main question is now, is it strictly necessary to have $\Set$ as codomain of this functors? The entire work is developed in a more general perspective, using the least possible concept directly connected with set theory, using instead categorial constructions. The property of $\Set$ we have made large use of is the equivalence between quotient and product subobjects, which holds in exact categories.
Hence, this work can be viewed as a starting point to a more general study on e-graphs, in which edges, nodes and equivalence classes can be not only sets but objects of any exact category.

>>>>

\iffalse
In questo caso, una introduzione meno brusca, ricapitolando un po' la struttura e soprattutto il fine della tesi

@@

The result we ended up with is then $\Reg(\EGG))$-adhesivity, where $\EGG$ is a category of functors which have $\Set$ as codomain category.
The main question is now, is it strictly necessary to have $\Set$ as codomain of this functors? The entire work is developed in a more general perspective, using the least possible concept directly connected with set theory, using instead categorial constructions. The property of $\Set$ we have made large use of is the equivalence between quotient and product subobjects, which holds in exact categories.
Hence, this work can be viewed as a starting point to a more general study on e-graphs, in which edges, nodes and equivalence classes can be not only sets but objects of any exact category.
\fi

Retracing the work, we have defined graphs with equivalence and e-graphs in a more abstract way, untied from potential implementations, in order to analyze formal properties of this kind of structures.

Then, based upon this framework, we proved adhesive properties of graphs with equivalences and e-graphs, obtaining an interesting result of adhesivity
(both of theses structure are adhesive with respect to the class of their regular monomorphisms)
useful for a potential exploiting in state-of-the-art techniques, such as the DPO approach for concurrent updates to the structure, currently not explored yet.

Both graphs with equivalence and e-graphs are considered in this work as functors with $\Set$ as codomain category.
One may ask then, is it strictly necessary to have $\Set$ as codomain of this functors?
The entire work is developed in a more general perspective, using the least possible concept directly connected with set theory, using instead categorial construction, and requiring only $\mathcal{M}$-adhesivity and exactness (this last concept is quite implicit in $\Set$, hence is not explicitly mentioned, but is, intuitively, the one that permitted the equivalence noted in \Cref{rem:eqgrph_set_eq}).
Hence, a future work will be a more general presentation of this structures, substituting $\Set$ with any category with this properties.

### 2024-11-22 Article/Graph_sections/Eq_Graphs_integration.tex
minima

\begin{remark}
     A graph with equivalence can be viewed as a graph endowed with an equivalence relation over its set of vertices, $(\graph G, \sim_\graph{G})$. An homomorphism between two graphs with equivalences $h :\eqgraph{G} = (\graph{G}, \sim_\graph{G})\rightarrow \eqgraph{H} = (\graph{H}, \sim_\graph{H})$ is a graph homomorphism $h = (h_V, h_E):\graph{G} \rightarrow \graph{H}$ such that if $v_1 \sim_\graph{G} v_2$ then $h_V(v_1) \sim_\graph{H} h_V(v_2)$. In $\Set$, it is possible to formalize an equivalence relation $\sim$ over $X$ as a surjective function sending each element $x$ on its equivalence class $[x]_{\sim}$, and this justify our formalization via surjective functions (i.e., epimorphisms).
\end{remark}

@@

\begin{cor}\label{cor:regmono}Let $\eqgraph{G} = (E_G, V_G, C_G, s_G, t_G, q_G)$ and $\eqgraph{H} = (E_H, V_H, C_H, s_H, t_H, q_H)$ be two graphs with equivalences, and let $ h = (h_E, h_V, h_C): \eqgraph{G}\to \eqgraph{H} $ be a morphism of $\EqGrph$, then the following are equivalent:
	\begin{enumerate}
		\item $h$ is a regular mono;
		\item $h_E$, $h_V$, $h_C$ are all monos;
%		\item $h_E$ and $h_V$ are mono and for every $v, v'\in V_H$, $q_H(h_V(v))=q_H(h_V(v'))$ if and only if $q_G(v)=q_G(v')$.
            \item $h_E$ and $h_V$ are mono and $(K, \pi_1, \pi_2)$ is the kernel pair of $q_H \circ h_V$ if and only if $(K, \pi_1, \pi_2)$ is the kernel pair of $q_G$.
	\end{enumerate}
\end{cor}
\begin{proof}
	$1\Rightarrow 2.$ If $h$ is mono, from \Cref{cor:mono_in_EqGrph} we have that $h_E$ and $h_V$ are monos. To derive $h_C$ mono, suppose $f, g :  \eqgraph{H \to K}$, where $\eqgraph{K} = (E_K, V_K, C_K, s_K, t_K, q_K)$to be the arrows equalized by $h$. Then we have
    \begin{align*}
        f_C \circ h_C \circ q_G  &=  f_C \circ q_H \circ h_V \\
                                            &=  q_K \circ f_V \circ h_V \\
                                            &=  q_K \circ g_V \circ h_V \\
                                            &=  g_C \circ h_C \circ q_G
    \end{align*}
    since $q_G$ is epi, we have that $f_C \circ h_C = g_C \circ h_C$, hence $h_C$ is an equalizer for $f_C$ and $g_C$, thus a monomorphism.

@@

	\smallskip \noindent 
	$3\Rightarrow 1.$\todo{Esercizio} \color{green}{idea: force the comm. of the diagram on the last two components to obtain the two arrows that are equalized, and show that the condition in 3 is sufficient to conclude reg. mono} \color{black}
\end{proof}

\begin{remark}
    It is possible to restate the third point of the \Cref{cor:regmono}, by \Cref{ex:kernel_pairs_in_Set}, as 
    \begin{displayquote}
    \textit{$h_E$ and $h_V$ are mono and, for every $v, v'\in V_H$, $q_H(h_V(v))=q_H(h_V(v'))$ if and only if $q_G(v)=q_G(v')$}
    \end{displayquote}
    That is, a regular monomorphism in $\EqGrph$ is a morphism that reflects equivalences besides preserving them.
\end{remark}

>>>>

\begin{remark}\label{rem:eqgrph_set_eq}
     A graph with equivalence can be viewed as a graph endowed with an equivalence relation over its set of vertices, $(\graph G, \sim_\graph{G})$. An homomorphism between two graphs with equivalences $h :\eqgraph{G} = (\graph{G}, \sim_\graph{G})\rightarrow \eqgraph{H} = (\graph{H}, \sim_\graph{H})$ is a graph homomorphism $h = (h_V, h_E):\graph{G} \rightarrow \graph{H}$ such that if $v_1 \sim_\graph{G} v_2$ then $h_V(v_1) \sim_\graph{H} h_V(v_2)$. In $\Set$, it is possible to formalize an equivalence relation $\sim$ over $X$ as a surjective function sending each element $x$ on its equivalence class $[x]_{\sim}$, and this justify our formalization via surjective functions (i.e., epimorphisms).
\end{remark}

@@

\begin{cor}\label{cor:regmono}
	Let $\eqgraph{G} = (E_G, V_G, C_G, s_G, t_G, q_G)$ and $\eqgraph{H} = (E_H, V_H, C_H, s_H, t_H, q_H)$ be two graphs with equivalences,
	and let $ h = (h_E, h_V, h_C): \eqgraph{G}\to \eqgraph{H} $ be a regular monomorphism of $\EqGrph$, then $h_E$, $h_V$, $h_C$ are all monos.
\end{cor}
\begin{proof}
	If $h$ is mono, from \Cref{cor:mono_in_EqGrph} we have that $h_E$ and $h_V$ are monos. To derive $h_C$ mono, suppose $f, g :  \eqgraph{H \to K}$, where $\eqgraph{K} = (E_K, V_K, C_K, s_K, t_K, q_K)$to be the arrows equalized by $h$. Then we have
    \begin{align*}
        f_C \circ h_C \circ q_G  &=  f_C \circ q_H \circ h_V \\
                                            &=  q_K \circ f_V \circ h_V \\
                                            &=  q_K \circ g_V \circ h_V \\
                                            &=  g_C \circ h_C \circ q_G
    \end{align*}
    since $q_G$ is epi, we have that $f_C \circ h_C = g_C \circ h_C$, hence $h_C$ is an equalizer for $f_C$ and $g_C$, thus a monomorphism.
	\iffalse

@@

	\smallskip \noindent 
	$3\Rightarrow 1.$\todo{Esercizio} \color{green}{idea: force the comm. of the diagram on the last two components to obtain the two arrows that are equalized, and show that the condition in 3 is sufficient to conclude reg. mono} \color{black}
	\fi
\end{proof}

\begin{prop}
	Let $\eqgraph{G} = (E_G, V_G, C_G, s_G, t_G, q_G)$ and $\eqgraph{H} = (E_H, V_H, C_H, s_H, t_H, q_H)$ be two graphs with equivalences,
	and let $ h = (h_E, h_V, h_C): \eqgraph{G}\to \eqgraph{H} $ be a regular monomorphism of $\EqGrph$. Then, $h_E$ and $h_V$ are mono and $(K, \pi_1, \pi_2)$ is the kernel pair of $q_H \circ h_V$ if and only if $(K, \pi_1, \pi_2)$ is the kernel pair of $q_G$.
\end{prop}

\begin{proof}
	By \Cref{cor:regmono}, we have that $h_E$, $h_V$ and $h_E$ are all monos. Hence, by \Cref{cor:kermono}, $(K, \pi_1, \pi_2)$ is the kernel pair of $q_G$ if and only if it is the kernel pair also of $h_C \circ q_G$, since $h_C$ is mono by hypothesis. The thesis follows from $h_C \circ q_G = q_H \circ h_V$, and from the hypothesis of $h_E$ mono.

\begin{remark}
    It is possible to restate the last proposition, by \Cref{ex:kernel_pairs_in_Set}, as 
    \begin{displayquote}
    \textit{$h_E$ and $h_V$ are mono and, for every $v, v'\in V_H$, $q_H(h_V(v))=q_H(h_V(v'))$ if and only if $q_G(v)=q_G(v')$}
    \end{displayquote}
    That is, a regular monomorphism in $\EqGrph$ is a morphism that reflects equivalences besides preserving them.
\end{remark}

### 2024-11-22 Article/introduction.tex
minima

@@

The developlment of equality graphs (e-graphs) is to be brought back to the eighties, in the context of automatic theorem provers (ATPs). E-Graphs are then an extension of prior structures, in order to represemt in a more compact way equivalence classes of expressions mantaining closure-under-congruence invariant.
One of the aims of this work is to present a more theoretic representation of e-graphs than the one at the state of the art presented in \cite{egg}, using category theory formalism, in order to integrate the existence of such structure in a research line concerning concurrency of rewrinig-based formalisms.
Indeed, graph-like structures turned out to be very advantageous for distributed system respresentation.
An example of this fact could be a graph as the abstraction a set of entities and relations between them within a state, while rewriting steps model computation steps that, modifying the graph result in changes of the state of the system.
A standard approach for this is the DPO (double pushout) approach, due to its flexibility and locality properties.
A presentation of e-graphs as the one in this work allows to move beyond the operational representation via data structures, in favour of a formalization that model the changes of the structure too.
In this setup, adhesivity results permit to make use of all the advantages on DPO rewriting, because of the generality of such framework.
The elegance of such formalism lies on the definition itself of adhesive categories, namely categories where it is guaranteed the existence of pullbacks and pushouts along (some) monomorphisms, which are additionally required to be Van Kampen (that is, stability of pushouts under pullbacks and stability of pullbacks under combined pushouts and pullbacks \cite{Ehrig_Golas}).
The only downside of such formalism is the difficulty of proofs using the definition. For cerain calasses of caeogries, there exists a charaterization of such property, presented in \cite{castelnovo2022newcriterionmathcalmmathcalnadhesivity}.

%Category theory provides uniform proofs and constructions across a wide range
%of models. The usual approach is to find a natural class of categories with the right
%structure to support the range of constructions particular to the application area.

>>>>

%\fi

@@

The developlment of visual formalisms such as equality graphs (e-graphs) is to be brought back to the eighties, in the context of automatic theorem provers (ATPs).
E-Graphs are then an extension of prior structures, in order to represent in a more compact way equivalence classes of expressions mantaining closure-under-congruence invariant.
Graph manipulation turns out to be very useful in practical fields such as systems optimization or even components upgrades, hence the development of formal frameworks is necessary, especially for what concernes parallelism and concurrency, as the research line in which this work wants to be integrated.

In recent years, indeed, graph-like structures found a wide diffusion as models to rewriting-based formalisms and dependency analisys.
An example of this fact could be a graph as the abstraction a set of entities and relations between them within a state, while rewriting steps model computation steps that, modifying the graph result in changes of the state of the system.

Due to its flexibility and locality properties, DPO (double pushout) approach and adhesive categrìories have become a standard model for this kind of considerations. 
A presentation of e-graphs as the one in this work allows to move beyond the operational representation via data structures, in favour of a formalization that model the changes of the structure too.
In this setup, adhesivity results permit to make use of all the advantages on DPO rewriting, because of the generality of such framework.
The elegance of such formalism lies on the definition itself of adhesive categories, namely categories where it is guaranteed the existence of pullbacks and pushouts along (some) monomorphisms, which are additionally required to be Van Kampen (that is, stability of pushouts under pullbacks and stability of pullbacks under combined pushouts and pullbacks \cite{Ehrig_Golas}).
The only downside of such formalism is the difficulty of proofs using the definition. For cerain calasses of caeogries, there exists a charaterization of such property, presented in \cite{castelnovo2022newcriterionmathcalmmathcalnadhesivity}.

Among all graphic frameworks, e-graphs, widely used in modern SMT (satisfiability modulo theories) solvers, such as Z3, are having success, due efficiency of their implementations, but is still missing a solid theoretical underpinning, since almost all the progress remains implementation-based.
Thus, one of the aims of this work is to present a more theoretic representation of e-graphs than the one at the state of the art presented in \cite{egg}, using categorial tools to be formalized.
Then, the main result is to show that e-graphs relapse in the current adhesive paradigm.
